Approximation as Differentiation

### 2025-07-09 main/language.tex
language subsection -> section.
Cleanup.
Tweaks.
Reinstate missing semantics.
[ tex ] fiddling
Just use + for binary coprod

\subsection{Higher-order language}

We introduce a total higher-order functional language, over a set $\PrimTy$ of primitive types $\rho$ and sets
$\PrimOp^\rho_{\rho_1,\ldots,\rho_n}$ of primitive operations $\phi$, equipped with sums, products, functions
and a data type of lists. We omit a treatment of general inductive or coinductive types.

@@

% \input{fig/positivity} -- follow nunes2023 instead
\input{fig/syntax}
% \input{fig/kinding}
\input{fig/typing}

\figrefTwo{syntax}{typing} give the syntax and typing rules.

@@

% \input{fig/mu-polynomial}
% \input{fig/inductive-types/semantics}
% \figref{semantics:types} gives the interpretation of types $\Delta \vdash \tau$ with $\length{\Delta}
% = n$ as functors $\Sem^n \to \Sem$, or as objects of $\Sem$ in the case $n = 0$.
% \todo{Notation for partially applied functor used in \figref{semantics:types}}
%
% \begin{definition}[$\mu$-polynomial]
% Suppose $\cat{C}$ a category with finite coproducts $(\coprod, 0)$ and finite products $(\times, 1)$. Define
% $\muPoly_{\cat{C}}$ to be the smallest subcategory of $\Cat$ generated inductively by the rules in
% \figref{mu-polynomial}.
% \end{definition}
%
% For any $\mu$-polynomial endofunctor $F: \cat{C} \to \cat{C}$ with initial algebra $\mu F$, write $\inMap_F$
% for the structure map $F\mu F \to \mu F$ and write $\cata_f$ for the unique morphism $\mu F \to X$ from the
% initial $F$-algebra to any $F$-algebra $(X, f: FX \to X)$.

@@

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(\coprod, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

>>>>

\section{Higher-Order Language}
\label{sec:language}

To model Galois slicing semantically for higher-order programs, we define a simple total functional
programming language, extending the simply-typed lambda calculus.

@@

The language includes base types $\rho$ drawn from a set $\PrimTy$, along with standard type formers for sums,
products, functions and lists, and an additional lifting type constructor $\tyLift$ to allow approximation
points to be added explicitly, as discussed in \secref{models-of-total-gps}. Terms include variables, the
usual introduction and elimination forms, a monadic return and bind for lifted terms, and primitive operations
$\phi$ of arity $n$ drawn from a family of sets $\PrimOp^\rho_{\rho_1,\ldots,\rho_n}$.

The language is intentionally minimal: it excludes general recursion, and general inductive or coinductive
types, which will consider in future work (\secref{conclusion}). Typing judgments for terms are standard and
shown in \figref{typing}, with the usual rules for products, sums, functions, lists and lifting.

\input{fig/syntax}
\input{fig/typing}

@@

@@

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(+, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}
